---
title: "Data tidying"
author: "Pascale"
date: "11/6/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
```


# read and clean data
```{r}
catch_original <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"),
                    stringsAsFactors = FALSE)
head(catch_original)
```

remove `all` column and `notes` column becasue we don't need them.
```{r}
catch_data <- catch_original %>% 
  #select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
  select(-All, -notesRegCode)

head(catch_data)
```


```{r}
summary(catch_data)
```

## fix chinook column

use `mutate` to fix Chinook column
```{r}
catch_clean <- catch_data %>% 
  #mutate(Chinook = as.numeric(Chinook))
  mutate(Chinook = ifelse(Chinook =="I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook))
```


finding row that was turned into NA
```{r}
i <- which(is.na(catch_clean$Chinook))
i
```

print row that contains a character rather than a number
```{r}
catch_data[i,]
```

# reshape data

```{r}
catch_long<-catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "count")

head(catch_long)
```

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = Year,
              values_from = count)

head(catch_wide)
```

skip this one
```{r}
#catch_long <- catch_long %>% 
  #rename(catch_thousands=count) %>% 
  #mutate(catch = catch_thousands * 1000) 

#head(catch_long)
```

```{r}
catch_long <- catch_long %>% 
  mutate(catch = count * 1000) %>% 
  select(-count)

head(catch_long)
```

# summarize data

```{r}
mean_region <- catch_long %>% 
  group_by(Region, species) %>% 
  summarise(catch_mean = mean(catch), 
            num_obs =n())

mean_region
```

```{r}
sum_sp <- catch_long %>% 
  group_by(species) %>% 
  summarise(sum_catch = sum(catch))

sum_sp
```

```{r}
total_sp_pre_1900 <- catch_long %>% 
  filter(Year < 1900) %>% 
  group_by(species) %>% 
  summarise(total = sum(catch)) %>% 
  arrange(desc(total))

total_sp_pre_1900
```

# join to region table

```{r}
region_defs <- read.csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.303.1",
                            method = "libcurl"),
                        stringsAsFactors = FALSE) %>% 
    select(code, mgmtArea) 
    # %>% rename(Region = code)


head(region_defs)
```

```{r}
catch_joined <- left_join(catch_long, region_defs, by = c("Region" = "code"))

#catch_joined <- left_join(catch_long, region_defs)

head(catch_joined)
```
functions that will help join datasets when characters are not identical:
* trimws - trips white space    
* tolower - changes to lower case    
* gsub - replace expressions (- vs. _)

# separate and unite
```{r}
site_codes <- data.frame(site = c("HAW-100",
                                  "HAW-101",
                                  "OAH-102",
                                  "OAH-103",
                                  "MAI-100"),
                         stringsAsFactors = F)
```

separate island and site number
```{r}
site_codes_split<- site_codes %>% 
  separate(site, into = c("island", "site_num"), sep="-")

site_codes_split
```

```{r}
site_codes_unite<- site_codes_split %>%
  unite(site.code, island, site_num, sep="_")

site_codes_unite
```

